Phylogeographic inference with microtellite and SNPs

### 2019-07-10 code/rmarkdown/Figures.rmd
subsample trees

```

>>>>

```
##Figure 2:

Assessment of optimal K values using (A-C) Pr[X|K] (Pritchard et al. 2000), (D-F) ΔK (Evanno et al. 2005) and (G-I) the parsimony index, PI (Wang 2019), from K = 1 to K = 8, for *H. molleri* (A, D, G), *P. cultripes* (B, E, H) and their subsampled datasets from n loci (n in 200, 500, 1000, 3000, 5000, 10000, 20000) (C, F, I). Panels A, B, D, E, G and H, depict results from SNP (red) and microsatellite (green) data, whereas panels C, F and I, depict results from SNP data alone.??I need to overplot SDs, change the line and pch pattern for usat/snp and color palette for subsets??

##Figure 3

Relation of bootstrap support to Dnr (distance from node to root) and Dnp (distance from node to parent node), for SNP (A, B, E, F) and microtellite (C, D, G, H) data from *H. molleri* (A, C, E, G) and *P. cultripes* (B, D, F, H).??should I overplot here the results from loci subsampling??

##Figure 4

Rooted neighbor joining trees for SNP (A, B) and microtellite (C-D) datasets from *H. molleri* (A, C) and *P. cultripes* (B, D). Nodes with bootstrap support above 0.70 are represented with black diamonds, or else with red dots. In each tip, pie chart depicts the ancestry proportions from CLUMPAK consensus solutions for the major cluster for K = 4 together with the population ID from Table 1.

### 2019-07-10 code/rmarkdown/article.Rmd
subsample trees

## 2.3 Population structure
We conducted model-based genetic structure analyses in STRUCTURE v2.3.4 (Pritchard et al. 2000). For each dataset, we performed 10 replicate runs from K = 1 to K = 8, since the best number of clusters fall within this range for both species (Gutiérrez-Rodríguez et al. 2017; Sánchez-Montes et al. 2019). We used an admixture model with correlated allele frequencies (Falush et al. 2003) and no prior information of sample origin. For the microtellite data, we used the same run lengths as in their original publications: 5*10^5 burnin followed by 1*10^6 iterations. For the SNP datasets, the run lengths were shorter, a burnin of 3*10^4 steps followed by 1 * 10^4 iterations, as for SNP data chain often converge fast (Pritchard, google groups...?). For the SNP data, we estimated lambda for K = 1, by averaging lambda estimates across 3 replicate runs. The values of lambda (0.60 for *H. molleri* and 0.69 for *P. cultripes*) were then used across all runs for the SNP data, whereas for microsatellites lambda was fixed to 1. These lower values of lambda can improve the modeling of correlated allele frequencies when using SNPs, where often the data is skewed toward rare alleles (Falush et al. 2003). STRUCTURE was run in parallel in 8 cores using Structure_threader (Pina-Martins et al. 2017), recording steps to log files every 50 and 5000 iterations for the SNP and microsatellite data, respectively. We run STRUCTURE analysis with random subsets of loci (n = 200, 500, 1000, 3000, 5000, 10000 and 20000) from the original matrix of SNPs for *P. cultripes* and the same STRUCTURE parameters. This approximation allowed to assess the effect of the number of loci in the patterns of genetic structure.
The convergence between the 10 replicate runs for each K was evaluated in R using Gelman and Rubin's convergence diagnostic (GR diagnostic; Gelman and Rubin 1992), with function *gelman.diag* in package *coda* (Plummer et al. 2006).
A central question in population genetics is determining the optimal number of populations subjacent to the genetic structure (Allendorf and Luikart 2007). However, hierarchical population structure (Evanno et al. 2005), unbalanced sampling and number markers (Wang 2017) can lead to the incorrect inference of number of populations. We used KFinder (Wang 2019) to compare the best number of populations for each dataset through 3 approaches: (1) Pr[X|K], the probability of data X given K populations (Pritchard et al. 2000) and, (2) ΔK, the most popular method and which considers the rate of change in the log probability of data between successive K (Evanno et al. 2005) and 3) PI, parsimony index, a newly proposed metric which favors K's which yield most consistent and with minimal average individual admixture, and which claims to be more consistent across wider demographic scenarios (Wang 2019).

@@

We evaluated the phylogenetic relationships between the populations using the neighbor-joining (NJ) algorithm (Saitou and Nei 1987). We discarded using model-based phylogenetic inference for the implicit difficulties of fitting evolutionary and demographic models for intra-species trees (), and for the possibilities that offer comparing the 2 marker types (SNPs and microtellites) under a similar tree-reconstruction method. The application of NJ relies on a distance matrix computed from the genotypes. For the microtellite datasets we computed these matrices using Da distance (Nei et al. 1983) with *genet.dist* function from *hierfstat* 0.04-22 package (Goudet and Jombart 2015) in R. Da distances have been shown to yield more reliable topologies to other distances (Takezaki and Nei 1996, 2008). For the SNP genotypes, we computed euclidean distances with *stats::dist* function in R. Then NJ trees were reconstructed with *njs* function from *ape* 5.0 (Paradis and Schliep 2018) for each dataset. We computed the bootstrap support with a modified approach from *ape::boot.phylo* to accommodate non-standard input data. First, we resampled with replacement loci in the genotype matrices, from which we recomputed NJ trees. These trees were rooted using midpoint rooting with function *midpoint.root* from *phytools* 0.6-60 (Revell 2012) in R. And finally, bootstrap values were computed by counting the partitions of the original trees present in 1000 bootstrapped trees.
*Marker type and bootstrap support*. For each tree, branch lengths were normalized by dividing edge lengths by the length between the most distant tips in the tree. Then, we used several functions in *treeman* 1.1.3 package (Bennett 2018) in R, to extract 2 metrics from each node in the trees: 1) the distance from each node to the root (Dnr), and 2) the distance from each node to its parent node (Dnp). ??For the nodes in each tree we assessed the relation of Dnr and Dnp against their corresponding bootstrap support values fitting generalized linear models (GMLs) using a logit link function, with *stats::glm* function in R. *Actually the fit is really bad. See script 99_04_model_fitting.R and ask Jeff*??

@@

# 3. RESULTS

After filtering (Figure S1), the dataset kept for analysis included a total `r totalind` individuals from `r nrow(samples)` localities: `r indH` individuals of *Hyla molleri* (microtellites: n = `r indHm`, `r locHm` loci, `r missHm` % missing data; SNPs: n = `r indHs`, `r locHs` loci, `r missHs` % missing data), from `r popH` localities, and `r indP` individuals of *Pelobates cultripes* (microtellites: n = `r indPm`, `r locPm` loci, `r missPm` % missing data; SNPs: n = `r indPs`, `r locPs` loci, `r missPs` % missing data), from `r popP` localities (Figure 1; Table S1).

## 3.1 Model-based clustering
*Convergence*. Convergence of the chains measured by the GR diagnostic varied between species and markers. Often, values below 1.05 indicate good convergence (Vats and Knudson 2018 Jeff should check this ref). For SNPs, the chains from each K in STRUCTURE converged well up to K = 2 in *H. molleri* and up to K = 6 in the case of *P. cultripes*. Reducing the number of loci did not affect the pattern of convergence across K in a consistent manner (Figures S2.1, S2.2, S2.5 - S2.11; Figure S3.1).
For microsatellite data, the chains converged consistently up to K = 6 in both species, although the GR diagnostic slightly peaked for K = 5 in *P. cultripes* (Figures S2.3 - S2.4; Figure S3.1).
*Optimal K*. The most optimal number of populations was evaluated with 3 metrics. Pr[X|K] increased steadily from K = 2 to K = 8 in all datasets (Figure 2 A-C). Evanno's ΔK peaked at K = 2 in all SNP datasets, followed by K = 3 (Figure 2 D-F). However, in the microsatellite dataset from *H. molleri*, ΔK peaked at K = 4 and K = 6 (Figure 2D). The predominant K = 2 in the SNP datasets were not in agreement with results from the Parsimony index (Figure 2 G-I), except for the microtellite data in *H. molleri*, in which PI and ΔK peaked at K = 4 (Figure 2G). For SNPs in *H. molleri*, the highest values of the PI were at K = 4 and K = 5, contrary to the K = 2 from ΔK (Figure 2G). In *P. cultripes*, the PI peaked at K = 4 for both markers, and then sunk in the SNP data, whereas for  microtellites the PI reached its maxima at K = 6 - K = 8 (Figure 2H). In the subsets of SNPs from *P. cultripes* the PI increased gradually up to K = 4. After that, the distinct marker sets showed inconsistent trends (Figure 2I).

## 3.2 Phylogenetic structure

*Coherence between model-based and NJ clustering*.

*Marker type and bootstrap support*.

>>>>

## 2.3 Population structure
We conducted model-based genetic structure analyses in STRUCTURE v2.3.4 (Pritchard et al. 2000). For each dataset, we performed 10 replicate runs from K = 1 to K = 8, since the best number of clusters fall within this range for both species (Gutiérrez-Rodríguez et al. 2017; Sánchez-Montes et al. 2019). We used an admixture model with correlated allele frequencies (Falush et al. 2003) and no prior information of sample origin. For the microtellite data, we used the same run lengths as in their original publications: 5*10^5 burnin followed by 1*10^6 iterations. For the SNP datasets, the run lengths were shorter, a burnin of 3*10^4 steps followed by 1 * 10^4 iterations, as for SNP data chain often converge fast (Pritchard, google groups...?). For the SNP data, we estimated lambda for K = 1, by averaging lambda estimates across 3 replicate runs. The values of lambda (0.60 for *H. molleri* and 0.69 for *P. cultripes*) were then used across all runs for the SNP data, whereas for microsatellites lambda was fixed to 1. These lower values of lambda can improve the modeling of correlated allele frequencies when using SNPs, where often the data is skewed toward rare alleles (Falush et al. 2003). STRUCTURE was run in parallel in 8 cores using Structure_threader (Pina-Martins et al. 2017), recording steps to log files every 50 and 5000 iterations for the SNP and microsatellite data, respectively. We run STRUCTURE analysis with random subsets of loci (n = 200, 500, 1000, 3000, 5000, 10000 and 20000) from the original matrix of SNPs for *P. cultripes* and the same STRUCTURE parameters. This approximation allowed to assess the effect of the number of loci in the patterns of genetic structure.
The convergence between the 10 replicate runs for each K was evaluated in R using Gelman and Rubin's convergence diagnostic (GR diagnostic; Gelman and Rubin 1992), with function *gelman.diag* in package *coda* (Plummer et al. 2006).
A central question in population genetics is determining the optimal number of populations subjacent to the genetic structure (Allendorf and Luikart 2007). However, hierarchical population structure (Evanno et al. 2005), unbalanced sampling and number markers (Wang 2017) can lead to the incorrect inference of number of populations. We used KFinder (Wang 2019) to compare the best number of populations for each dataset through 3 approaches: (1) Pr[X|K], the probability of data X given K populations (Pritchard et al. 2000) and, (2) ΔK, the most popular method and which considers the rate of change in the log probability of data between successive K (Evanno et al. 2005) and 3) PI, parsimony index, a newly proposed metric which favors K's which yield most consistent and with minimal average individual admixture, and which claims to be more consistent across wider demographic scenarios (Wang 2019).
We ran CLUMPAK (Kopelman et al. 2015) on STRUCTURE outputs. CLUMPAK feeds the software CLUMPP with the replicate runs at each value of K to generate consensus solutions for the distinct modes. It also computes the similarity between ancestry matrices from each run and it matches clusters across successive values of K.

@@

We evaluated the phylogenetic relationships between the populations using the neighbor-joining (NJ) algorithm (Saitou and Nei 1987). We discarded using model-based phylogenetic inference for the implicit difficulties of fitting evolutionary and demographic models for intra-species trees (), and for the possibilities that offer comparing the 2 marker types (SNPs and microtellites) under a similar tree-reconstruction method. The application of NJ relies on a distance matrix computed from the genotypes. For the microtellite datasets we computed these matrices using Da distance (Nei et al. 1983) with *genet.dist* function from *hierfstat* 0.04-22 package (Goudet and Jombart 2015) in R. Da distances have been shown to yield more reliable topologies to other distances (Takezaki and Nei 1996, 2008). For the SNP genotypes, we computed euclidean distances with *stats::dist* function in R. Then NJ trees were reconstructed with *njs* function from *ape* 5.0 (Paradis and Schliep 2018) for each dataset. We computed the bootstrap support with a modified approach from *ape::boot.phylo* to accommodate non-standard input data. First, we resampled with replacement loci in the genotype matrices, from which we recomputed NJ trees. These trees were rooted using midpoint rooting with function *midpoint.root* from *phytools* 0.6-60 (Revell 2012) in R. And finally, bootstrap values were computed by counting the partitions of the original trees present in 1000 bootstrapped trees. We inspected the phylogeographic congruence between STRUCTURE and the phylogenetic by plotting together a the consensus solution for the major cluster from the output of CLUMPAK for K = 4, since K = 4, seemed to be an optimal solution for many of the datasets (Figure 2).
*Marker type and bootstrap support*. For each tree, branch lengths were normalized by dividing edge lengths by the length between the most distant tips in the tree. Then, we used several functions in *treeman* 1.1.3 package (Bennett 2018) in R, to extract 2 metrics from each node in the trees: 1) the distance from each node to the root (Dnr), and 2) the distance from each node to its parent node (Dnp). ??For the nodes in each tree we assessed the relation of Dnr and Dnp against their corresponding bootstrap support values fitting generalized linear models (GMLs) using a logit link function, with *stats::glm* function in R. *Actually the fit is really bad. See script 99_04_model_fitting.R and ask Jeff*??

@@

# 3. RESULTS
After filtering (Figure S1), the dataset kept for analysis included a total `r totalind` individuals from `r nrow(samples)` localities: `r indH` individuals of *Hyla molleri* (microtellites: n = `r indHm`, `r locHm` loci, `r missHm` % missing data; SNPs: n = `r indHs`, `r locHs` loci, `r missHs` % missing data), from `r popH` localities, and `r indP` individuals of *Pelobates cultripes* (microtellites: n = `r indPm`, `r locPm` loci, `r missPm` % missing data; SNPs: n = `r indPs`, `r locPs` loci, `r missPs` % missing data), from `r popP` localities (Figure 1; Table S1).

## 3.1 Model-based clustering
*Convergence*. Convergence of the chains measured by the GR diagnostic varied between species and markers. Often, values below 1.05 indicate good convergence (Vats and Knudson 2018 Jeff should check this ref). For SNPs, the chains from each K in STRUCTURE converged well up to K = 2 in *H. molleri* and up to K = 6 in the case of *P. cultripes*. Reducing the number of loci did not affect the pattern of convergence across K in a consistent manner (Figures S2.1, S2.2, S2.5 - S2.11; Figure S3.1).
For microsatellite data, the chains converged consistently up to K = 6 in both species, although the GR diagnostic slightly peaked for K = 5 in *P. cultripes* (Figures S2.3 - S2.4; Figure S3.1).
*Optimal K*. The most optimal number of populations was evaluated with 3 metrics. Pr[X|K] increased steadily from K = 2 to K = 8 in all datasets (Figure 2 A-C). Evanno's ΔK peaked at K = 2 in all SNP datasets, followed by K = 3 (Figure 2 D-F). However, in the microsatellite dataset from *H. molleri*, ΔK peaked at K = 4 and K = 6 (Figure 2D). The predominant K = 2 in the SNP datasets were not in agreement with results from the Parsimony index (Figure 2 G-I), except for the microtellite data in *H. molleri*, in which PI and ΔK peaked at K = 4 (Figure 2G). For SNPs in *H. molleri*, the highest values of the PI were at K = 4 and K = 5, contrary to the K = 2 from ΔK (Figure 2G). In *P. cultripes*, the PI peaked at K = 4 for both markers, and then sunk in the SNP data, whereas for  microtellites the PI reached its maxima at K = 6 - K = 8 (Figure 2H). In the subsets of SNPs from *P. cultripes* the PI increased gradually up to K = 4. After that, the distinct marker sets showed inconsistent trends (Figure 2I).
*clumpak table*

## 3.2 Phylogenetic structure
*Coherence between model-based and NJ clustering*. The phylogeographic structure of the model-based ancestries for the SNP dataset were coherent with the topology of their phylogenetic trees (Figure 4 A-B). That is, the samples from the same populations clustered together on the same or close clades and the ancestries from the samples were coherent with the phylogenetic relationships at different depths in the tree. However, the topology of the trees for microtellites were little supported. Furthermore, it neither provided a solid grouping of the samples from the same populations and it was not able to explain the relationships at different depths in the trees (Figure 4 C-D).

*Marker type and bootstrap support*. The topology of the trees from SNP data were highly supported compared to the trees from microtellites (Figures 3 and 4). The overall support was higher for the SNP dataset of *P. cultripes* than for *H. molleri* (Figure 3 A-B; Figure A-B). In microtellite data, larger distances of a node to the root (Dnr) were strongly correlated with higher bootstrap support (??statistics??)(Figure 3 C-D; Figure 4 C-D). This correlation was ¿¿not?? significant for the SNP data (??statistics??)(Figure 3 A-B; Figure 4 A-B). The distance of a node its precedent in direction to the root, or parent node (Dnp), explained a huge variation of in the bootstrap support for all datasets ??statistics?? (Figure 3 E-I; Figure 4 C-D). In the case of the SNP data Dnp ranged from 0? to 0.15?, but almost all trees supported nodes that had a Dnp above 0.025?. The slope and R^2 of the logarithmic regression for the SNP data were very large compared to the microsatellites (??statistics??; Figure 3 E-I). In *P. cultripes*, where the number of SNPs in the dataset was larger than in *H. molleri*, the slope was greater (statistics??; Figure 3 E-F)
